---
title: "Predicting diabetes risk"
format:
  html:
    code-fold: true
execute:
  echo: true
author: "Hamza Rahal"
date: "2024-10-17"
categories: [project, code, analysis]
---

# **Logistic Regression in R -** Predicting diabetes risk

## Definition

Logistic regression is a statistical analysis method to predict a binary outcome, such as yes or no, based on prior observations of a data set.
A logistic regression model predicts a dependent data variable by analyzing the relationship between one or more existing independent variables. For example, in health, Logistic regression can also be used in the following areas:
• To identify risk factors and plan preventive measures;
• In drug research to tease apart the effectiveness of medicines on health outcomes across age, gender
and ethnicity;

## About the project

I used data of 318 individuals with and without diabetes type 2.The aim is to examine the relationship
between age, gender, BMI, diet type, smoking status and family history of disease and build a model for
predicting diabetes risk

### Importing data into R and loading packages

```{r}
Diabetes <- read.csv("diabetes.csv")
if(!require(pacman)) install.packages("pacman")
## Loading required package: pacman
pacman::p_load( tidyverse, janitor, plotrix, gtsummary, survival )
```

### Data manipulation and exploration

```{r}
Diabetes$gender <- factor(Diabetes$gender, levels = c("1","2"),
labels = c("male","female"))
Diabetes$smoking <- factor(Diabetes$smoking, levels = c("1","2"),
labels = c("smokers","Non-smokers"))
Diabetes$diabetes <- factor(Diabetes$diabetes, levels = c("0","1"),
labels = c("Non-diabetic","Diabetic"))
Diabetes$veg <- factor(Diabetes$veg, levels = c("1","2"),
labels = c("vegeterian","Non-vegeterian"))
Diabetes$familiy_history <- factor(Diabetes$familiy_history, levels = c("1","2"),
labels = c("Yes","No"))

Diabetes <- Diabetes %>%
mutate(BMI = weight/heightˆ2)
Diabetes <- Diabetes %>%
select(age, gender, smoking,weight, diabetes,veg, familiy_history, BMI) %>%
mutate(Agegroup=
ifelse(age < 65, "< 65", "> =65"),
BMIclass = case_when(
BMI < 18.5~ "Underweight",
BMI >=18.5 & BMI < 24.9 ~ "Normal weight",
TRUE ~ "Obese"))
dim(Diabetes)

```

```{r}
table1 <- table(Diabetes$diabetes)
piepercent <-paste0(round(100 * table1/sum(table1)), "%")
plotrix::pie3D(table1,radius = 1.0,
explode = 0.05,
labels = piepercent,
col = c("purple","orange"),
main = "Diabetic Vs Non-diabetic compositions",
2
col.main="blue")
legend("topright",
c("Non-diabetic","diabetic"),
cex = 0.8,
fill =c("purple","orange") )
```

### Descriptive statistics

```{r}
descriptives <- Diabetes %>%
select(age,BMI,diabetes,Agegroup,gender,BMIclass,smoking,veg,familiy_history)
tbl_summary(descriptives,
type = list(familiy_history ~ "categorical") ) %>%
modify_caption("Table1: Descriptive statistics") %>%
bold_labels()
```

### Testing for significance between diabetic and non diabetic

```{r}
Diabetes %>%
select(diabetes,age,BMI,Agegroup,gender,BMIclass,smoking,familiy_history,veg) %>%
tbl_summary(by=diabetes,
label = list(age ~ "mean age,yrs (sd)",
BMI~ "median BMI (IQR)",
smoking ~ " smoking status",
veg ~ " Diet type",
familiy_history ~ " Family history"),
statistic = list(age ~ "{mean}({sd})"),
type = list(familiy_history ~ "categorical")) %>%
bold_labels() %>%
add_p(test = list(
all_continuous() ~ "t.test",
all_categorical() ~ "fisher.test")) %>%
modify_caption("Table2: Statistical significance difference by diabetes status")
```

### Univariate analysis of diabetes risk factors

```{r}
Diabetes %>%
select(diabetes,Agegroup,gender,BMIclass,smoking,familiy_history,veg) %>%
tbl_uvregression(
method = glm,
y = diabetes,
method.args = list(family = binomial),
exponentiate = TRUE) %>%
modify_caption("Table3: Univariate analysis of diabetes risk factors") %>%
bold_p() %>%
italicize_levels() %>%
bold_labels()
```
